### 2021-07-08 rad-icfp-2021/correctness.tex
added a paragraph 7.2
some progress correctness

A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
free multicategy on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.

@@

    \begin{tabular}{r c l}
    $\seml \reals \semr$ & $\defeq$& $\RR$ \\
    $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ \\
    !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ 
    \end{tabular}
    \begin{tabular}{r c l} 
        !$\seml$op2$\semr$! & $\defeq$ & op2: $\reals\times \reals\to\reals$ \\
        !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & TODO \\
        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ &
    \end{tabular}

Variables are interpreted as projections, let binding uses composition in the multicategory. 
Pairs are interpreted using the Cartesian structure of the multicategory.
\end{definition}

We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
However, Target is higher-order but $\CartSp$ is not Cartesian Closed. 
Instead, we can for instance interpret Target in the category of Diffeological spaces, 
as in \cite{huot2020correctness}. 
Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
The key idea will be that a higher order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
to build first order functions that are not smooth.

@@

A diffeological space is thus a set equipped with structure.
Many constructions of sets carry over straightforwardly to diffeological spaces.

\begin{example}[Cartesian diffeologies]\label{ex:cartesian-diffeologies}
Each open subset $U$ of $\RR^n$ can be given the structure of a diffeological space by taking all the
smooth functions $V\to U$ as $\plots{U}^V$.
It is easily seen that smooth functions from $V\to U$ in the traditional sense coincide with
smooth functions in the sense of diffeological spaces.
Thus diffeological spaces have a profound relationship with ordinary calculus.
\end{example}

\begin{example}[Product diffeologies]
Given a family $\seq[i\in I]{X_i}$ of diffeological spaces,
we can equip the product $\prod_{i\in I}X_i$ of sets with the
\emph{product diffeology} in which $U$-plots are precisely the functions
of the form $\seq[i\in I]{p_i}$ for $p_i\in\plots{X_i}^U$.  
\end{example}

\begin{example}[Functional diffeology]
We can equip the set $\Diff(X,Y)$ of smooth functions between diffeological spaces with the \emph{functional diffeology}
in which $U$-plots consist of functions $f:U\to \Diff(X,Y)$ such that 
$(u,x)\mapsto f(u)(x)$ is an element of $\Diff(U\times X, Y)$.
\end{example}

We can thus interpret function types !$\seml$A -> B$\semr$! = $\Diff$($\seml$!A!$\semr$,$\seml$!B!$\semr$).

@@

In particular, one can form a syntactic concategory on some base types and primitives, 
which satisfies a similar universal property as syntactic multicategories.
Two partiular examples are of interest for us, as they will allow us to interpret Source and Target UNF
but they will also help us explain the UNF and $UNF^{-1}$ transformations.

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ be the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$, 
TODO: $map2_{x,y.e}$and $reduce_{x,y.e1;e2}$.
\end{definition}

One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.

\begin{notation}[Morphisms in a Cartesian concategory]
    TODO: projection, unique map, $\concatcomp$, swap, pair, id, ...
\end{notation}

We can interpret Source UNF in $\ConcatS$ as follows.

\begin{tabular}{l c l}
   $\seml$!var!$_{T;i} \semr$ &=& $<id_\Gamma,\pi_i>$ \\
   $\seml$!op!$_{T;n} \semr$ &=& $id_\Gamma\concatcomp op_n$\\
   $\seml$!pair!$_{T;A\times B} \semr$ &=& $pair_{T;A\times B}$ \\
   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp 1_{T2}\concatcomp id_{T3}$\\
   $\seml$!e1!$\comp$!e2!$\semr$  &=& $\seml$!e1!$\semr \comp \seml$!e2!$\semr$ \\
   $\seml$!e1!$\pcomp$!e2!$\semr$ &=& $\seml$!e1!$\semr \comp (\seml$!e2!$\semr\concatcomp id )\comp swap$ \\
   $\seml$!map2!$_{T;x,y.e}\semr$  &=& $id_\Gamma\concatcomp map2_{x,y.e}$ \\
   $\seml$!reduce!$_{T;x,y.e;e}\semr$ &=& $id_\Gamma\concatcomp reduce_{x,y.e1;e2}$ \\
\end{tabular}

@@

TODO: I think need to explain here the (co)monoid structure induced by $\Dsynrevsymbol_{\rho}$

@@

\begin{definition}[Mutlicat from concat]
    We define $\CSource$ to be the multicategory with the same objects as Source and 
    $\CSource([A_1,\ldots,A_n],B)=\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B])$.
\end{definition}

@@

\begin{definition}[Concat from multicat]
    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
    with morphisms $A\to [B_1,\ldots,B_n]$ being $n$ morphisms $A\to Bi$ of $\catC$. 
\end{definition}

\begin{proposition}[Initiality of concat2 gives $UNF^-1$]
    TODO
\end{proposition}

@@

\begin{definition}[D as a lax functor]
    TODO
\end{definition}

@@

Combining the the previous propositions, we have shown that $\directD{\rho}{\Gamma}{Y}$ is correct.

>>>>

A term $x_1:A_1,\ldots,x_n:A_n\vdash e:A$ is interpreted as a morphism $\sem{e}:[\sem{A_1},\ldots,\sem{A_n}]\to \sem{A}$ in a multicategory.
Substitution is interpreted as composition. We first consider a syntactic model for a language, which consists of a 
free multicategory on some base types and primitives. Our source and target languages induce syntactic multicategories as follows.

@@

    \begin{tabular}{r c l}
    $\seml \reals \semr$ & $\defeq$& $\RR$ \\
    $\seml$!T1xT2!$\semr$ & $\defeq$& $\seml$!T1!$\semr \times\seml$!T2!$\semr$ \\
    $\seml \Array{\reals}{n}\semr$ & $\defeq$ & $\prod_{1\leq i \leq n} \RR$ \\
    !$\seml$op1$\semr$! & $\defeq$ & op1: $\reals\to\reals$ 
    \end{tabular}
    \begin{tabular}{r c l} 
        !$\seml$op2$\semr$! & $\defeq$ & op2: $\reals\times \reals\to\reals$ \\
        !$\seml$c$\semr$! & $\defeq$ & $c\in\RR$ \\
        !$\seml$map2 (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$ ! & $\defeq$ & !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>;swap;$(\seml$ e$_1\semr)^{\times n}$! \\
        !$\seml$reduce (x,y.e$_1$) e$_2$ e$_3\semr_\gamma$! & $\defeq$ <> &
    \end{tabular}
where $\Delta_n$ is the $n$-copy map !<id,$\ldots$,id>!.
The semantics for !reduce (x,y.e$_1$) e$_2$ e$_3$! is given by first !<$\Delta_n,$,$\seml$ e$_2\semr$,$\seml$ e$_2\semr$>!, following a swap isomorphism  $\Gamma^n \times \RR \times \RR^n \to \reals \times (\reals \times \Gamma)^n$. 
Finally, we apply $\seml$ e$_1\semr \times id$ $n$ times, where the identity is of the obvious type at each stage.

Variables are interpreted as projections $\pi_i$, let binding as composition in the multicategory. 
Pairs are interpreted using the Cartesian structure of the multicategory.
\end{definition}

We can similarly interpret Target in a multicategory of smooth-like spaces and functions. 
However, Target is higher-order and $\CartSp$ is not Cartesian Closed. 
Instead, we can interpret Target in the category of Diffeological spaces, 
as in \cite{huot2020correctness}. 
Diffeological spaces (\cite{iglesias2013diffeology}) are a conservative extension of $\CartSp$. 
The key idea will be that a higher-order function is called smooth if it sends smooth functions to smooth functions, meaning that we can never use it
to build first order functions that are not smooth.

@@

A diffeological space is thus a set equipped with structure.
Many constructions of sets carry over straightforwardly to diffeological spaces.
For instance, given a family $\seq[i\in I]{X_i}$ of diffeological spaces,
we can equip the product $\prod_{i\in I}X_i$ of sets with the
\emph{product diffeology} in which $U$-plots are precisely the functions
of the form $\seq[i\in I]{p_i}$ for $p_i\in\plots{X_i}^U$.  
Cartesian spaces $\RR^n$ can be given the structure of a diffeological space by taking all the
smooth functions $U\to \RR^n$ as $\plots{\RR^n}^V$. We can equip the set $\Diff(X,Y)$ of smooth functions between diffeological spaces with the \emph{functional diffeology}
in which $U$-plots consist of functions $f:U\to \Diff(X,Y)$ such that 
$(u,x)\mapsto f(u)(x)$ is an element of $\Diff(U\times X, Y)$.
We can thus interpret function types !$\seml$A -> B$\semr$! = $\Diff$($\seml$!A!$\semr$,$\seml$!B!$\semr$).

@@

In particular, one can form a syntactic concategory on some base types and primitives, 
which satisfies a similar universal property as syntactic multicategories.
Two particular examples are of interest for us, as they will allow us to interpret Source and Target UNF
but they will also help us explain the UNF and $UNF^{-1}$ transformations.

\begin{definition}[$\ConcatS$]
Let $\ConcatS$ be the syntactic Cartesian concategory whose types are those of Source and 
with primitives given by $op1: \reals \to \reals$, $op2: \reals,\reals \to \reals$, 
$map2_{x,y.e}: \Array{\reals}{n},\Array{\reals}{n}\to \Array{\reals}{n}$ and $reduce_{x,y.e1;e2}:\Array{\reals}{n}\to \reals$.
\end{definition}

One may notice that the syntax of $\ConcatS$ is somewhere in between the syntax of Source and of Source UNF.
Given two morphisms !e$_1$:T$_1$->T$_2$! and !e$_2$:T$_3$->T$_4$!, we denote by !e$_1 \concatcomp$ e$_2$:T$_1$,T$_3$->T$_2$,T$_4$!.
We denote by $u_T$ the unique morphism from !T! to the terminal object $[]$.
We can interpret Source UNF in $\ConcatS$ as follows.

\begin{tabular}{l c l}
   $\seml$!var!$_{T;i} \semr$ &=& $<id_\Gamma,\pi_i>$ \\
   $\seml$!op!$_{T;n} \semr$ &=& $id_\Gamma\concatcomp op_n$\\
   $\seml$!pair!$_{T;A\times B} \semr$ &=& $pair_{T;A\times B}$ \\
   $\seml$!proj!$_{T1;T2;T3} \semr$ &=& $id_{T1}\concatcomp u_{T2}\concatcomp id_{T3}$\\
   $\seml$!e1!$\comp$!e2!$\semr$  &=& $\seml$!e1!$\semr \comp \seml$!e2!$\semr$ \\
   $\seml$!e1!$\pcomp$!e2!$\semr$ &=& $\seml$!e1!$\semr \comp (\seml$!e2!$\semr\concatcomp id )\comp swap$ \\
   $\seml$!map2!$_{T;x,y.e}\semr$  &=& $id_\Gamma\concatcomp map2_{x,y.e}$ \\
   $\seml$!reduce!$_{T;x,y.e;e}\semr$ &=& $id_\Gamma\concatcomp reduce_{x,y.e1;e2}$ \\
\end{tabular}

@@

@@

\begin{definition}[Mutlicat from concat]
    We define $\CSource$ to be the multicategory with the same objects as Source and 
    $\CSource([A_1,\ldots,A_n],B)=\{f\in\ConcatS([A_1,\ldots,A_n],[A_1,\ldots,A_n,B], \forall i.f;\pi_{A_i}=id_{A_i})$.
    The composition of $f_i:\underline{A}_i\to B_i$ with $g:\underline{B}\to C$ is given 
    $f_1\concatcomp\ldots f_n;(u_{\underline{B}_1}\concatcomp A_1 \concatcomp \ldots \concatcomp u_{\underline{B}_n}\concatcomp A_n);g$.
    In other words each term $f_i$ forgets about its output $\underline{B}_i$, and then we use the composition in the concategory.
\end{definition}

We can interpret Source in $\CSource$ as follows.
TODO

@@

\begin{definition}[Concat from multicat]
    A multicategory $\catC$ naturally defines a concategory with the same objects as $\catC$ and
    with morphisms $\underline{A}\to [B_1,\ldots,B_n]$ being $n$ morphisms $\underline{A}\to B_i$ of $\catC$. 
\end{definition}

This allows to form a concategory $\CTarget$ from the syntactic multicategory for Target.
We can interpret $\ConcatT$ in $\CTarget$ as follows.
TODO

This interpretation is in essence $\UNFSymbol^{-1}$. 
The difference is that to preserve typing, the semantic $\UNFSymbol^{-1}$
sends a non Jacobian primitive to a tuple, as in the example \ref{}. 
This is highly inefficient and the the syntactic $\UNFSymbol^{-1}$ 
additionally projects to the last element, where the non trivial information of the term is.

@@

\begin{definition}[$\Dsynrevsymbol_{\rho}$ as a lax functor]
    We define $\Dsynrevsymbol_{\rho}:\ConcatS\to\ConcatT$ as follows.
    $\Dsynrevsymbol_{\rho}$(![A$_1$,$\ldots$,A$_n$])=[A$_1$,$\ldots$,A$_n$,A$_1\times\ldots\times$A$_n\to \rho$]!.
\end{definition}

@@

Combining the previous propositions, we have shown that $\directD{\rho}{\Gamma}{Y}$ is correct.
